---
title: "Example_PLSR_Analysis"
author: "JStark"
date: "`r Sys.Date()`"
output: html_document
---

## Code Information

Code to run Partial Least Squares Regression Analysis
Code adopted from from Dr. H. Abdi
Modified by Jessica H. Stark, M.S., The Ohio State University

## Download packages needed for analysis
```{r}
library(corrplot)
library(ExPosition)
library(tidyverse)
library(PTCA4CATA)
library(data4PCCAR) # PLSR 
library(boot)
library(dplyr)
```

## Read in your X-matrix and Y- Matrix
These should be cleaned datasheets with your predictor (x) and outcome (y) variables of interest
Exclude subject ID from these CSVs since that will not be one of your variables
```{r}
X_matrix <- read.csv('EXAMPLE.csv')  #CSV file of all of your predictor variables (in other words, your X-matrix)

Y_matrix <- read.csv('EXAMPLE.csv')  #CSV file of all of your outcome variables (in other words, your Y-matrix)
```

## Partal Least Squares Regression (PLSR) Setup 
```{r}
set.seed(3)  #Set the seed so that you get the same result every tim

#Use function "normaliz" to center both dataframes (required for PLSR)
X_matrix_normal <- normaliz(X_matrix)
Y_matrix_normal <- normaliz(Y_matrix)
```

## Compute Partial Least Squares Regression (PLSR)
```{r}
#Uses "PLSR_SVD" function
#Start with number of total possible components (number of columns in X_matrix), then explore screeplot to see how many components to analyze
#This is best implemented with training and then test data 

resultPLSR <- PLSR_SVD(X_matrix_normal[1:nrow(X_matrix_normal),], Y_matrix_normal[1:nrow(Y_matrix_normal),], ncol(X_matrix_normal), #number of columns in X_matrix 
                    inference = TRUE,
                    displayJack = FALSE)
```

## Explore Scree Plots
```{r pressure, echo=FALSE}
#Scree Plot for the X-Matrix 
#Uses function "PlotScree"

Scree_plot_x <- PlotScree(ev    = resultPLSR$R2x, #percent variance explained per PLSR component, x-matrix 
   title = 'PLSR: X-Matrix Scree Plot',
)
```

```{r}
# Pseudo scree Y
Scree_plot_y <- PlotScree(ev    = resultPLSR$R2y, #percent variance explained per PLSR component, y-matrix 
   title = 'PLSR: Y-Matrix Scree Plot',
)
```

##Run PLSR with optimal number of components 
```{r}
resultPLSR_2 <- PLSR_SVD(X_matrix_normal[1:nrow(X_matrix_normal),], Y_matrix_normal[1:nrow(Y_matrix_normal),], 2, #this number will change based on the analysis of the scree plot
                    inference = TRUE,
                    displayJack = FALSE)
```

## Explore circle of correlation for components 1 and 2, X-matrix
```{r}
#Use function "correlationPlotter"
#This can then be modified for other components of the PLSR model
X_matrix_correlation_plot <- correlationPlotter(data_matrix = X_matrix_normal,
                   x_axis = 1, #component 1
                   y_axis = 2, #component 2 
                   factor_scores = resultPLSR_2$T #where factor scores are saved in the results matrix
                   )
```

##Explore circle of correlation components 1 and 2, y-matrix
```{r}
#Same as above, but for Y-matrix variables
Y_matrix_corr_plot <- correlationPlotter(data_matrix = Y_matrix_normal,
                   x_axis = 1, #component 1
                   y_axis = 2, #component 2
                   factor_scores = resultPLSR_2$T) #where factor scores are saved in the results matrix
```

##Explore results by a given categorical variable

###Import categorical data 
```{r}
Categorical_data <- read.csv('CATEGORICAL_DATA.csv')
```

###The example below is for Age Group
```{r}
#Color age groups for graphing purposes
color_age_group <- as.factor(Categorical_data$Age_Group)
age_group_color <- recode(color_age_group, 
          'Young_Adult' = 'blue',
          'Middle_Aged' = 'green',
          'Older_Adult' = 'deeppink') 
```

###Graph the observations by age group, for components 1 and 2 
```{r}
#Below is a map which plots each participant and how they load onto components 1 and 2, colored by age group
participant_map_age_groups <- PTCA4CATA::createFactorMap(
  resultPLSR_2$T, 
  title = 'Participant Loadings on 
  Components 1 and 2 by Age Group',
  col.points = age_group_color,
  alpha.points = .3,
  pch = 19,
  display.labels = FALSE,
  text.cex = 12,
  col.axes = 'black',
  col.background = 'white',
)

#Create labels ----
label_x_axis_participant_map <- labs(x = paste0('Component ',1,'. Variance Explained = ',
                            round(resultPLSR_2$R2y[1],2) ))
label_y_axis_participant_map <- labs(y = paste0('Component ',2,'. Variance Explained = ',
                            round(resultPLSR_2$R2y[2],2) )) 

#Now add the labels and make the text bigger on the original participant map 
participant_map_age_group_labeled <- participant_map_age_groups$zeMap + label_x_axis_participant_map + label_y_axis_participant_map + theme(text = element_text(size = 14)) #change text size to 14

#Print the map 
print(participant_map_age_group_labeled)
```

### Create bootstrapped means per component per age group, 1000 bootstraps
```{r}
#Use "Boot4Mean" function to bootstrap the factor scores from the PLSR results to assess how a given group loads on a given component 
boot_means_age_group <- Boot4Mean(Data=resultPLSR_2$T,design=color_age_group,niter=1000)
```

### Calculate bootstrap ratios
These bootstrap ratios tell us whether a GROUP is significantly loading on a given PLSR-derived component
```{r}
#Use function "boot.ratio.test" to obtain bootstrap ratios and look at $BootCube results to explore by group 
bootstrap_ratio_age_group <- boot.ratio.test(boot_means_age_group$BootCube)
```

###Now add ellipses to the graph to represent the 95% confidence interval for each bootstrapped group mean 
Below code obtained from from Dr. Ju-chi Yu 
```{r}
#Re-create the coloring variable for plotting
age_group_color_index <- c('Young_Adult' = 'blue',
                           'Middle_Aged' = 'green',
                            'Older_Adult' = 'deeppink')

#Make BootCube output it's own variable (same as bootcube variable from boot4mean above)
boot_cube_age_group <- boot_means_age_group$BootCube

#Rename these columns so to say "factor" in front of them for clarity
colnames(boot_cube_age_group) <- paste0("Factor ", c(1:ncol(resPLSR_2$T)))

#Create ellipses using "MakeCIELLIPSES" function
Bootstrapped_ellipses_age_group <- MakeCIEllipses(boot_cube_age_group,
                                   names.of.factors = paste0("Factor ", c(1:2)),
                                   col = age_group_color_index[rownames(boot_cube_age_group)],
                                   alpha.ellipse = 0.1,
                                   line.size = 0.8, alpha.line = 0.8) #these are parameters for your elipses, change per your preference 
```

###Create a map that just has the bootstrapped group means
```{r}
participant_map_age_group_confidence_intervals <- PTCA4CATA:: createFactorMap(
  boot_means_age_group$BootstrappedGroupMeans,
  col.points = age_group_color_index[rownames(boot_means_age_group$BootstrappedGroupMeans)],
  alpha.points = 1,
  pch = 19,
  cex = 4
)
```

###Now using all of the individual maps you just created, create a graph that has individual participant data, bootstrapped group means, and 95% confidence interval ellipses
```{r}
participant_map_age_groups_full <- participant_map_age_groups$zeMap_background  + participant_map_age_groups$zeMap_dots + label_x_axis_participant_map + label_y_axis_participant_map + Bootstrapped_ellipses_age_group + participant_map_age_group_confidence_intervals$zeMap_dots + theme(text = element_text(size = 14))

print(participant_map_age_groups_full)
```
